Temperature and moisture effects on greenhouse gas emissions from deep active-layer

### 2016-04-22 manuscript/cpcrw_manuscript.Rmd
Abstract written; intro tweaked; typos fixed

@@

High latitude ecosystems are being subjected to rapid climatic changes, with rising air temperatures and fires expected to drive permafrost degradation and soil carbon (C) cycle changes. How these soils' greenhouse gas emissions will respond to changes in temperature, moisture, and vegetation is highly uncertain. We used a laboratory experiment to examine how temperature and moisture control CO~2~ and CH~4~ emissions from soils sampled from the bottom of the annual active layer in an Alaskan boreal forest. Gas emissions were measured from thirty cores, subjected to two temperatures and either field moisture conditions or drought, over a 100-day incubation. TODO...

@@

High latitude ecosystems are being subjected to rapid changes in climate {IPCC, 2013 #3916} and fire occurrence and intensity {Kasischke, 2010 #2631}. This has a wide variety of effects on ecosystems and society, notably in northwestern North America and Alaska {Hinzman, 2005 #4315}{Ju, 2016 #4320}. In particular, rising temperatures and increasing fire will result in permafrost degradation {Pastick, 2015 #4311}{Zhang, 2015 #4372} and changes in soil temperature and hydrology that are expected to increase soil greenhouse gas (GHG) fluxes to the atmosphere, promoting a positive climate feedback {Treat, 2015 #4217}{Koven, 2011 #3092}.

The magnitude, timing, and form-in particular as methane (CH~4~) or carbon dioxide (CO~2~)-of such a feedback remain highly uncertain {Schuur, 2015 #4319}. While northern soils hold enormous quantities {Tarnocai, 2009 #2231} of potentially mineralizable soil organic carbon (SOC), vegetation succession dynamics promote permafrost resilience to even large temperature changes {Jorgenson, 2010 #2627}. Such dynamics may however be disrupted by increased fire disturbance {Johnstone, 2010 #2465}. In addition, the stability of SOC is itself highly uncertain, as it depends on soil temperature and moisture, the ages of and ratio between the carbon (C) and nitrogen (N) pools {Weiss, 2015 #4317}, and its protection from competent microorganisms, enzymes, and resources {Bailey, 2012 #3624}, whether by organomineral sorption, chemical lability, or physical location {Schmidt, 2011 #3135}.

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. A long literature of laboratory incubations, field observations, and meta-analyses has documented increased CO2 fluxes with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. GHG responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw are expected to lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, and interact with soil temperature effects in long-term incubations {Elberling, 2013 #3720}. More uniformly across the landscape, however, decadal warming, and perhaps drying, trends in Alaska {Bieniek, 2014 #4389} will work to counteract these effects, with uncertain consequences for the carbon cycle and climate system.

The goal of this study was to examine how temperature and moisture control C loss and GHG (CO~2~ and CH~4~) emissions from soils sampled from the bottom of the annual active layer--i.e., directly above permafrost--in an Alaskan boreal forest. Most previous studies have focused on surface soils or permafrost soils, neglecting deep active-layer soils that experience fluctuating temperature and, especially, moisture conditions. Such deep but unfrozen soils were identified as subject to strong effects from a two-decade warming experiment in the Alaskan Arctic {Sistla, 2013 #3666}. We also aimed to characterize the chemical and biological properties of these soils as they were subjected to a 100-day incubation at different temperatures, subjecting some cores to drying treatments. We hypothesized that (i) CO~2~ would be the dominant pathway for C loss in these largely aerobic soils; (ii) soils maintained at field moisture and high temperature would lose more C-CO~2~ than cores incubated at 4˚C, due to increased aerobic and anaerobic microbial activity; and (iii) core CH~4~ fluxes would be sensitive only to temperature, as no anaerobic conditions were planned for the cores.

@@

In the lab, the soil cores were stored at 4 °C for several days until they were weighed and prepared for incubation. At that point (11-12 August 2015), three fragmented or otherwise damaged cores were discarded, and the remaining cores were randomly assigned to one of six groups (N=6 in each group). These groups included two incubation temperatures (4 and 20 °C), each with two moisture treatments: one in which soil moisture was maintained at field conditions (~28% moisture by volume), and a drought treatment in which no water was added and cores were allowed to dry down to ~5% moisture by volume. The fifth group was a 20 °C "controlled drought" one, in which water was added so that these cores' moisture status would close match those of the 4 °C "drought" cores, which we anticipated would dry more slowly than their 20 °C counterparts. The final 6-core group was used for destructive, pre-incubation measurements including moisture content, pH, soil carbon and nitrogen, and bulk density. Subsamples were collected and stored at -20 °C for dissolved organic carbon measurements.

@@

Cumulative respiration for each core and gas was calculated by linearly interpolating flux rates between measurement dates and summing respired C over the entire incubation. The effect of temperature and treatment (drought, controlled drought, or field moisture conditions) on cumulative gas fluxes was evaluted with a post-hoc Tukey Honest Significant Differences test. Temperature sensitivity (Q~10~) was calculated for each gas and treatment as $\frac{F_2}{F_1}^{\left( \frac{10}{T_2-T_1} \right)}$ where *F~1~* and *F~2~* are the cumulative gas fluxes (mg C) at temperatures *T~1~* and *T~2~* (°C), respectively.

@@

```{r wc, echo=FALSE}
# wc and wc_alltrt are computed by 5-analyze.R script
gwcbegin <- subset(gwc_alltrt, stage == "Beginning")
gwcend <- subset(gwc, stage == "End")
gwcendmean <- gwcend$mean
names(gwcendmean) <- gwcend$Treatment
gwcendsd <- gwcend$sd
names(gwcendsd) <- gwcend$Treatment
vwcbegin <- subset(vwc_alltrt, stage == "Beginning")
vwcend <- subset(vwc_alltrt, stage == "End")
```

@@

After removal of known problematic data, and exclusion of statistical outliers, CO~2~ flux rates measured during the incubation ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CO2"], 1)` µg C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CO2"], 1)` µg C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CO2"], 1)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CO2"], 2)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CO2"], 1)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CO2"], 1)`). CH~4~ rates ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CH4"]*1000, 2)` ng C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CH4"]*1000, 2)` ng C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CH4"]*1000, 2)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CH4"]*1000, 3)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CH4"]*1000, 2)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CH4"]*1000, 2)`).

These means conceal considerable variablility over the course of the incubation (**Figures 2 and 3**). In the linear mixed-effects model, CO~2~ was strongly influenced by incubation chamber temperature, core gravimetric water content, and percent soil N (all P < 0.05, and the latter two P < 0.001; **Table 1**). (Note that percent C and percent N were highly correlated (r ≥ 0.98) for these cores. Because percent N was a slightlyl stronger predictor, it was retained in the model while percent C was excluded.) The interaction between water content and soil N was also highly significant (`r pclean(summary(step_co2_lme)$tTable["WC_gravimetric:N_percent", "p-value"])`), with high-N cores having little relationship between water content and CO~2~ flux (data not shown). "Drought" cores' cumulative production was `r pn(drought_effect["4", "Reduction"], 0)`% (4 °C) and `r pn(drought_effect["20", "Reduction"], 0)`% (20 °C) lower than the cores kept at field moisture.

@@

```{r compute_ratios, echo=FALSE}
# Ratio data is stored in the `gas_ratio` d.f. computed by analysis script
minmag <- min(gas_ratio$CO2CH4_ratio) %>% log10 %>% floor %>%
  c("one", "two", "three", "four", "five", "six", "seven", "eight", "nine")[.]
minratio <- min(gas_ratio$CO2CH4_ratio / 1e6)  %>% formatC(digits = 1, format = 'f')
mintrt <- gas_ratio$Treatment[which.min(gas_ratio$CO2CH4_ratio)]
mintemp <- gas_ratio$Temperature[which.min(gas_ratio$CO2CH4_ratio)]
maxratio <- max(gas_ratio$CO2CH4_ratio / 1e6) %>% formatC(digits = 1, format = 'f')
maxtrt <- gas_ratio$Treatment[which.max(gas_ratio$CO2CH4_ratio)]
maxtemp <- gas_ratio$Temperature[which.max(gas_ratio$CO2CH4_ratio)]
```

The cumulative production of C from CO~2~ was over `r minmag` order of magnitudes higher than that from CH~4~, with CO~2~:CH~4~ C ratios ranging from `r minratio` million in the `r mintemp` °C "`r maxtrt`" treatment, to `r maxratio` million in the `r maxtemp` °C "`r maxtrt`" treatment. Cumulative CO~2~ evolved was highly affected by temperature (`r pclean(co2_hsd$Temperature["p adj"])`), and "field moisture" cores emitted significantly more CO~2~ than the other two treatments at both temperatures (P < 0.001 for both). There was no difference between the 20 °C "drought" and "controlled drought" treatments (`r pclean(co2_hsd$Treatment["Drought-Controlled drought", "p adj"])`). Neither temperature (`r pclean(ch4_hsd$Temperature[,"p adj"])`) nor treatment (mean `r pclean(mean(ch4_hsd$Treatment[,"p adj"]))`) was a significant factor in predicting cumulative CH~4~ fluxes.

@@

Warming usually increases soil GHG fluxes, for example at depth in a long-term Arctic tundra experiment {Sistla, 2013 #3666}, as increased temperatures increment enzyme kinetics, enhance the production of extracellular enzymes, and enhance desorptions rates of organic matter from minerals. A key question, for both experimentalists and modelers {Falloon, 2011 #3045}, is to what degree such soils' emissions could by constrained by their moisture status--in particular, limited by low soil moisture driven by increases in high-latitude temperatures, vapor pressure deficit, and potentially precipitation changes.

>>>>

```{r wc, echo=FALSE}
# Compute soil moisture data for use in abstract and results below
# wc and wc_alltrt are computed by 5-analyze.R script
gwcbegin <- subset(gwc_alltrt, stage == "Beginning")
gwcend <- subset(gwc, stage == "End")
gwcendmean <- gwcend$mean
names(gwcendmean) <- gwcend$Treatment
gwcendsd <- gwcend$sd
names(gwcendsd) <- gwcend$Treatment
vwcbegin <- subset(vwc_alltrt, stage == "Beginning")
vwcend <- subset(vwc_alltrt, stage == "End")
```

```{r compute_ratios, echo=FALSE}
# Compute ratios between CO2 and CH4 for use in abstract and results below
# Ratio data is stored in the `gas_ratio` d.f. computed by analysis script
minmag <- min(gas_ratio$CO2CH4_ratio) %>% log10 %>% floor %>%
  c("one", "two", "three", "four", "five", "six", "seven", "eight", "nine")[.]
minratio <- min(gas_ratio$CO2CH4_ratio / 1e6)  %>% formatC(digits = 1, format = 'f')
mintrt <- gas_ratio$Treatment[which.min(gas_ratio$CO2CH4_ratio)]
mintemp <- gas_ratio$Temperature[which.min(gas_ratio$CO2CH4_ratio)]
maxratio <- max(gas_ratio$CO2CH4_ratio / 1e6) %>% formatC(digits = 1, format = 'f')
maxtrt <- gas_ratio$Treatment[which.max(gas_ratio$CO2CH4_ratio)]
maxtemp <- gas_ratio$Temperature[which.max(gas_ratio$CO2CH4_ratio)]
```

@@

Rapid climatic changes, rising air temperatures, and increased fires are expected to drive permafrost degradation and soil carbon (C) cycle changes in many high-latitude ecosystems. How these soils will respond to changes in their temperature, moisture, and overlying vegetation is highly uncertain, but critical to understand given the large soil C stocks in these regions. We used a laboratory experiment to examine how temperature and moisture control CO~2~ and CH~4~ emissions from mineral soils sampled from the bottom of the annual active layer, i.e. directly above permafrost, in an Alaskan boreal forest. Gas emissions from thirty cores, subjected to two temperatures and either field moisture conditions or experimental drought, were tracked over a 100-day incubation; we also measured a variety of physical and chemical characteristics of the cores. Gravimetric water content was `r pn(gwcbegin$mean, 2)` ± `r pn(gwcbegin$sd, 2)` (unitless) at the beginning of the incubation; cores kept at field moisture were unchanged at the end, but drought cores had declined to `r pn(gwcendmean["Drought"], 2)` ±  `r pn(gwcendsd["Drought"], 2)`. Carbon dioxide fluxes were strongly influenced by incubation chamber temperature, core water content, and percent soil N. Methane fluxes were most strongly correlated with percent N, but neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes, however. The cumulative production of C from CO~2~ was over `r minmag` order of magnitudes higher than that from CH~4~. The CO~2~ temperature sensitivity (i.e. Q~10~) was `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)` for the field moisture and drought treatments, respectively. Deep but unfrozen high-latitude soils have been shown to be strongly affected by long-term experimental warming, and these results help us understand the potential future dynamics of such soils with future climate and permafrost changes.

@@

High latitude ecosystems are being subjected to rapid changes in climate {IPCC, 2013 #3916} and fire occurrence and intensity {Kasischke, 2010 #2631}, notably in northwestern North America and Alaska {Hinzman, 2005 #4315}{Ju, 2016 #4320}. This will have a wide variety of effects on both ecosystems and society: in particular, rising temperatures and increasing fire will likely result in permafrost degradation {Pastick, 2015 #4311}{Zhang, 2015 #4372} and changes in soil temperature, with subsequent hydrology changes that will influence soil greenhouse gas (GHG) fluxes to the atmosphere. Such fluxes are a large component of the global C cycle and could thus promote a positive climate feedback {Treat, 2015 #4217}{Koven, 2011 #3092}.

The magnitude, timing, and form-in particular as methane (CH~4~) or carbon dioxide (CO~2~)-of such any such feedback remain highly uncertain {Schuur, 2015 #4319}. While northern soils hold enormous quantities {Tarnocai, 2009 #2231} of potentially mineralizable soil organic carbon (SOC), vegetation succession dynamics promote permafrost resilience to even large temperature changes {Jorgenson, 2010 #2627}. Such dynamics may however be disrupted by increased fire disturbance {Johnstone, 2010 #2465}. In addition, the stability of SOC is itself highly uncertain, as it depends on soil temperature and moisture, the ages of and ratio between the carbon (C) and nitrogen (N) pools {Weiss, 2015 #4317}, and its protection from competent microorganisms, enzymes, and resources {Bailey, 2012 #3624}, whether by organomineral sorption, chemical lability, or physical location {Schmidt, 2011 #3135}.

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. A long literature of laboratory incubations, field observations, and meta-analyses has documented increased CO~2~ fluxes with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. GHG responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw are expected to lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}; such interactions are critical to examine in the course of long-term incubation experiments {Elberling, 2013 #3720}. More uniformly across the landscape, however, decadal warming, and perhaps drying, trends in Alaska {Bieniek, 2014 #4389} will work to counteract these effects, with uncertain consequences for the carbon cycle and climate system.

The goal of this study was to examine how temperature and moisture control C loss and GHG (CO~2~ and CH~4~) emissions from soils sampled from the bottom of the annual active layer–i.e., directly above permafrost–in an Alaskan boreal forest. Most previous studies have focused on surface soils or permafrost soils, neglecting deep active-layer soils that were identified as subject to strong effects from a two-decade warming experiment in the Alaskan Arctic {Sistla, 2013 #3666}. We also aimed to characterize the chemical and biological properties of these soils as they were subjected to a 100-day incubation at different temperatures, subjecting some cores to drying treatments. We hypothesized that (i) CO~2~ would be the dominant pathway for C loss in these largely aerobic soils; (ii) soils maintained at field moisture and high temperature would lose more C-CO~2~ than cores incubated at 4˚C, due to increased aerobic and anaerobic microbial activity; and (iii) core CH~4~ fluxes would be sensitive only to temperature, as no anaerobic conditions were planned for the cores.

@@

In the lab, the soil cores were stored at 4 °C for several days until they were weighed and prepared for incubation. At that point (11-12 August 2015), three fragmented or otherwise damaged cores were discarded, and the remaining cores were randomly assigned to one of six groups (N=6 in each group). These included two incubation temperatures (4 and 20 °C), each with two moisture treatments: one in which soil moisture was maintained at field conditions (~28% moisture by volume), and a drought treatment in which no water was added and cores were allowed to dry down to ~5% moisture by volume. The fifth group was a 20 °C "controlled drought" one, in which water was added so that these cores' moisture status would close match those of the 4 °C "drought" cores, which we anticipated would dry more slowly than their 20 °C counterparts. The final 6-core group was used for destructive, pre-incubation measurements including moisture content, pH, soil carbon and nitrogen, and bulk density. Subsamples were collected and stored at -20 °C for dissolved organic carbon measurements.

@@

Cumulative respiration for each core and gas was calculated by linearly interpolating flux rates between measurement dates and summing respired C over the entire incubation. The effect of temperature and treatment (drought, controlled drought, or field moisture conditions) on cumulative gas fluxes was evaluated with a post-hoc Tukey Honest Significant Differences test. Temperature sensitivity (Q~10~) was calculated for each gas and treatment as $\frac{F_2}{F_1}^{\left( \frac{10}{T_2-T_1} \right)}$ where *F~1~* and *F~2~* are the cumulative gas fluxes (mg C) at temperatures *T~1~* and *T~2~* (°C), respectively.

@@

@@

After removal of known problematic data, and exclusion of statistical outliers, CO~2~ flux rates measured during the incubation ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CO2"], 1)` µg C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CO2"], 1)` µg C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CO2"], 1)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CO2"], 2)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CO2"], 1)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CO2"], 1)`) over the 100 days. CH~4~ rates ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CH4"]*1000, 2)` ng C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CH4"]*1000, 2)` ng C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CH4"]*1000, 2)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CH4"]*1000, 3)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CH4"]*1000, 2)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CH4"]*1000, 2)`).

These means conceal considerable variablility over the course of the incubation (**Figures 2 and 3**). In the linear mixed-effects model, CO~2~ was strongly influenced by incubation chamber temperature, core gravimetric water content, and percent soil N (all P < 0.05, and the latter two P < 0.001; **Table 1**). (Note that percent C and percent N were highly correlated (r ≥ 0.98) for these cores. Because percent N was a slightly stronger predictor, it was retained in the model while percent C was excluded.) The interaction between water content and percent N was also highly significant (`r pclean(summary(step_co2_lme)$tTable["WC_gravimetric:N_percent", "p-value"])`), with high-N cores having little relationship between water content and CO~2~ flux (data not shown).

@@

The cumulative production of C from CO~2~ was over `r minmag` order of magnitudes higher than that from CH~4~, with CO~2~:CH~4~ C ratios ranging from `r minratio` million in the `r mintemp` °C "`r maxtrt`" treatment, to `r maxratio` million in the `r maxtemp` °C "`r maxtrt`" treatment. Cumulative CO~2~ evolved was highly affected by temperature (`r pclean(co2_hsd$Temperature["p adj"])`), and "field moisture" cores emitted significantly more CO~2~ than the other two treatments at both temperatures (P < 0.001 for both). There was no difference between the 20 °C "drought" and "controlled drought" treatments (`r pclean(co2_hsd$Treatment["Drought-Controlled drought", "p adj"])`). "Drought" cores' cumulative production was `r pn(drought_effect["4", "Reduction"], 0)`% (4 °C) and `r pn(drought_effect["20", "Reduction"], 0)`% (20 °C) lower than the cores kept at field moisture. Neither temperature (`r pclean(ch4_hsd$Temperature[,"p adj"])`) nor treatment (mean `r pclean(mean(ch4_hsd$Treatment[,"p adj"]))`) was a significant factor in predicting cumulative CH~4~ fluxes.

@@

Warming usually increases soil GHG fluxes, for example at depth in a long-term Arctic tundra experiment {Sistla, 2013 #3666}, as increased temperatures increment enzyme kinetics, enhance the production of extracellular enzymes, and enhance desorptions rates of organic matter from minerals. A key question, for both experimentalists and modelers {Falloon, 2011 #3045}, is to what degree such soils' emissions could by constrained by their moisture status–in particular, limited by low soil moisture driven by increases in high-latitude temperatures, vapor pressure deficit, and potentially precipitation changes.
